Subset coercions in \Coq \\ \small{Draft - \today}

### 2006-08-23 papers/article/main.tex
Rework example, various nice fixes

\newenvironment{code}{\begin{alltt}\vspace{-1em}}{\vspace{-1em}\end{alltt}}
% Caml2Tex

@@

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{head} which returns
the first element of a list. In \ML{}:
\begin{code}
  let head = function hd :: tl -> hd | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
  Definition head (l : list A) : option A :=
    match l with
      hd :: tl => Some hd
    | [] => None
    end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, using a
richer specification: 
\begin{code}
  Definition head (l : list A) : forall p : l <> [] -> A.
\end{code} 
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. We propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is a weak, \emph{decidable} type-checking
procedure along with an elaboration into partial \Coq terms. Using
\Russell, we would write:
\begin{code}
  Definition head \{ l : list A | l <> [] \} : A := 
  match l with
    hd :: tl -> hd
  | [] -> _
  end
\end{code}

The specification shows that we're defining a partial function yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proof in the second branch. After type-checking,
the proof assistant will ask the user to prove that in the context where
\id{l} is a non-empty list and $l = []$, we can create an object of type
$A$ (which should be obvious as the context is contradictory).

@@

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-1em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset}
  \vspace{-2.5em}
\end{figure}

@@

    \vspace{\infvspace}
    \Abs\DP

    \vspace{\infvspace}
    \App\DP

@@

\begin{figure}[h]
  \subtdRules
  \caption{\Russell{} conversion}
  \label{fig:conversion-decl-rules}
\end{figure}

@@

    \vspace{\infvspace}
    \PiLeftA\DP

    \vspace{\infvspace}
    \PiRightA\DP

@@

Finally, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U~s, `G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}
\begin{proof}
  Rules are syntax-directed.
\end{proof}

@@

Our interpretation will build a full-fledged \CIC term from its
algorithmic skeleton and a rich type written in \Russell. Obviously, we
cannot infer proof terms where they are needed in \CIC, but we can build
a \emph{partial} term, leaving typed holes where proofs are
needed. Hence we add a rule to build existential variables (or
metavariables) in the target calculus ($\typeq$ denotes the new system's typing judgment):
\begin{prooftree}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
\end{prooftree}

We restrict it to objects of type $\Prop$ because we consider that the
informative part of the algorithm has been entirely given in the
original term.

@@

\subtis[Coercion derivation]

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied. Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgment $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and well suited for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

@@

\interpfigs[Interpretation of terms]

@@

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms into a system with only
$\beta\pi$-equivalence. 
Indeed, if our algorithmic system has transitivity and symmetry of
coercions, we expect that the derived coercions will inherit these
properties, for example:
\begin{proposition}[Symmetry of coercions]
  If $\subimpl{`G}{c}{T}{U}{s}$ then there exists $c^{-1}$ so that
  $\subimpl{`G}{c^{-1}}{U}{T}{s}$ and $c `o c^{-1} `= c^{-1} `o c `=
  \ctxdot$.
\end{proposition}

Now take the following example:
\begin{example}[Symmetry for products]
  Let $T, U = \Pi x : A.B$. The coercions $c, c^{-1}$ derived by
  the algorithm must have the form $\lambda x :
  \ip{A}{`G}.c_2[\ctxdot~c_1[x]]$, hence we must include the $\eta$
  rule in order to have $c `o c^{-1} `= \ctxdot$.
\end{example}

Similarly, we must add $\eta$ rules for all introduction terms of our
language if we want to be conservative (the coercion algorithm will
always derive $\eta$-long normal forms, so we must be able to prove that they
are equivalent to the empty coercion) and have unicity of
coercions. The later is an important property of our system, as it
corresponds in practice to the determinism of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the calculus of constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.

\ccqeqfig[Equational theory of \CICq]

@@

\section{The \Program{} tactic}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system.

@@

\subsection{Related work}
Other efforts to build certified programs using type theory include
Cayenne \cite{Augustsson99} and Epigram
\cite{DBLP:journals/jfp/McBrideM04}. Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker, it can be sought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementally to get a complete program.
Using type annotations and with the help of the editor for structuring, 
they are able to write programs with precise specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scaling issues. DML \cite{Xi98:PhD} is more akin to
comparison with our solution. It provides a way to use a restricted set
of dependent types in \ML programs, carefully chosen so as to keep
type-checking decidable by addition of an automatic prover. Our method
subsumes this one as it is perfectly possible to integrate automatic
proof tools with \Coq to resolve the generated obligations.

>>>>

\newenvironment{code}{\begin{alltt}\vspace{-1.4em}}{\vspace{-1.4em}\end{alltt}}
% Caml2Tex

@@

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{tail} which returns
the rest element of a non-empty list. In \ML{}:
\begin{code}
  let tail = function hd :: tl -> tl | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
  Definition tail (l : list A) : option A :=
    match l with
      hd :: tl => Some tl
    | [] => None
    end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, and also
attach a property to the result to express more precisely what \id{tail}
does:
\begin{code}
  Definition tail (l : list A) : forall p : l <> [] -> 
  \{ l' : list A | exists a, a :: l' = l \}.
\end{code}
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. We propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is a weak, \emph{decidable} type-checking
procedure which does not require proofs to be present in the code when
constructing objects of a subset type. Using \Russell, we would write:
\begin{code}
  Definition tail \{ l : list A | l <> [] \} : 
    \{ l' : list A | exists a, a :: l' = l \} := 
  match l with
    hd :: tl -> tl
  | [] -> []
  end
\end{code}

The specification shows that we're defining a partial function and
shows the relation betwenn input and ouput, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proofs in both branches. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list $tl$ has the property \verb|exists a, a :: tl = hd :: tl|
  and
\item In the context where
  \id{l} is a non-empty list and $l = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

@@

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-2em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset}
  \vspace{-1em}
\end{figure}

@@

    \vspace{\infvspace}
    \Abs\DP
    \quad\hspace{-2.5em}
    \App\DP

@@

\begin{figure}[h]
  \vspace{-2em}
  \subtdRules
  \vspace{-2em}
  \caption{\Russell{} conversion}
  \label{fig:conversion-decl-rules}
  \vspace{-1em}
\end{figure}

@@

    \vspace{\infvspace}
    \PiLeftA\DP
    \quad\hspace{-2.5em}
    \PiRightA\DP

@@

Finally, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U~s, `G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}
\begin{proof}
  Rules are syntax-directed.
\end{proof}

@@

Our interpretation will build a full-fledged \CIC term from its
algorithmic skeleton and a rich type written in \Russell. Obviously, we
cannot infer proof terms where they are needed in \CIC, but we can build
a \emph{partial} term, leaving typed holes where proofs are
needed. Hence we add a rule to build existential variables (or
metavariables) in the target calculus ($\typeq$ denotes the new system's typing judgment):

\begin{center}
  \vspace{-1.7em}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
  \DP
  \vspace{-1.7em}
\end{center}
We restrict it to objects of type $\Prop$ because we consider that the
informative part of the algorithm has been entirely given in the
original term.

@@

\begin{figure}[ht]
  \vspace{-1em}
  \subtisc
  \vspace{-2em}
  \caption{Coercion derivation}
  \label{fig:coerce-impl-rules-short}
  \vspace{-1em}
\end{figure}

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgment $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and adapted for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

@@

\begin{figure}[ht]
  \vspace{-2em}
  \interparr
  \vspace{-2em}
  \caption{Interpretation of terms}
  \label{fig:interp}
  \vspace{-1em}
\end{figure}

@@

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms into a system with only
$\beta\pi$-equivalence and still be conservative with respect to \CIC. 
Indeed if we want the \irule{Conv} rule to be admissible, we must have for example:
If $\Pi x : A.B \eqbp \Pi x : C.D$ then there exists $c$ such that
$\subimpl{`G}{c}{\Pi x : A.B}{\Pi x : C.D}{s}$ and $c `= \ctxdot$.
However, we derive eta-long coercions, so $c$ will be $\lambda x :
\ip{C}{`G}.x$.

Similarly, we must add $\eta$ rules for all introduction terms of our
language if we want to be conservative and have unicity of
coercions. The later is an important property of our system, as it
corresponds in practice to the determinism of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the calculus of constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{-1em}
  \ccqeqarr
  \vspace{-2em}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{-1em}
\end{figure}

@@

\section{The \Program{} tactic}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The type-checker also transforms pattern-matching so
that an equality between the matched term and the pattern is present in
each branch typing context and it allows the user to put explicit holes in the term, for example:
\begin{code}
  Definition head \{ l : list A | l <> 0 \} : A :=
    match l with
    | hd :: tl -> hd
    | [] -> _
    end.
\end{code}
Here the second branch can't be taken because we would have $l = []$ and
$l \neq 0$, but we may not have any placeholder element of type $A$.

@@

\subsection{Related work}
Other efforts to build certified programs using type theory include
Ca\-yenne \cite{Augustsson99} and Epigram
\cite{DBLP:journals/jfp/McBrideM04}. Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker, it can be sought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementally to get a complete program.
Using type annotations and with the help of the editor for structuring, 
they are able to write programs with precise specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scaling issues. DML \cite{Xi98:PhD} is more akin to
comparison with our solution. It provides a way to use a restricted set
of dependent types in \ML programs, carefully chosen so as to keep
type-checking decidable by addition of an automatic prover. Our method
subsumes this one as it is perfectly possible to integrate automatic
proof tools with \Coq to resolve the generated obligations.
